Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-10-25 paper/paper.tex
example rough draft

% -*- fill-column: 80; -*-
\documentclass[sigconf,review,anonymous,screen,acmsmall]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{xspace}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{stmaryrd}
\usepackage{scalerel}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows, fit}

@@

\newcommand{\bang}{~\textbf{!}~}
\newcommand{\nil}{\mathsf{nil}}
\newcommand{\cons}{\mathsf{cons}}
\newcommand{\data}{\mathsf{data}}
\newcommand{\stringquote}[1]{\texttt{\textquotesingle#1\textquotesingle}}

@@

To gain intuiton for working in \theoryname, let's walk through some examples of
grammars and parses internal to it.

\paragraph{Finite Grammars}
First, consider finite grammars over the alphabet $\{ a , b , c \}$. Each letter of the
alphabet corresponds to a base type in our calculus. For instance, the grammar
for the single character $a$ has a single parse tree for the string \stringquote{a}.
and no parse trees at any other strings.

A finite grammar is built from base types via dijunctions $\oplus$ and
concatenations $\otimes$ --- for instance, $(a \oplus b) \otimes c$.
In \cref{fig:fingram} we find a proof tree showing that the input string \stringquote{ac} matches
the grammar $(a \oplus b) \otimes c$.

@@

\paragraph{Regular Expressions}
Next, we demonstrate how to parse a regular expression in \theoryabbv. Again,
consider grammars over $\{ a , b , c \}$. Similar to finite grammars, regular
expresssions are built inductively using base types, $\oplus$, and $\otimes$ ;
however, unlike finite grammars, regular expressions additionally have a
restricted form of recursion via Kleene star $(\cdot)^{\ast}$.

% TODO this should be beter formatted but can't use listings with \otimes ^{\ast}
\begin{figure}
\begin{align*}
\data &~A^{*}:\\
      & \nil : I \to A* \\
      & \cons : A \otimes A* \to A*
\end{align*}
\caption{Kleene Star as an inductive type}
\label{fig:kleenestarinductive}
\end{figure}

@@

However, we can have derivations where the term in context is not simply a
string of base types. For instance, by using the elimination principle for
Kleene star, \cref{fig:kleeneabstractproof} builds a term of type
$A^{*}$ in context comprising abstract\footnote{A linear type that is not a
  finite grammar.} linear types. In these contexts, it is not quite appropriate
to treat linear terms as parse \textit{trees}, because the information in
context is not a string. Instead, it is more
appropriate to think of such terms as parse \textit{transformers}. That is, in
\cref{fig:kleeneabstractproof} the
inhabitant $x : (A \otimes A)^{*}$ in the context is not itself data to be parsed; rather,
$x$ is a parse tree and the term from context $x : (A \otimes A)^{*}$ to the
term $A^{*}$ is a transformation of parse trees of the input grammar into parse
trees of the output grammar.

@@

% \begin{tikzpicture}[%opacity=0.5,
%   treenode/.style = {shape=rectangle, rounded corners,
%     draw, align=center,
%     top color=white, bottom color=white!20},
%     line/.style={draw, -latex'},
%     edge from parent/.style={draw,-latex'},
%     level 1/.style={sibling distance=20mm, level distance=0.5cm},
%     level 2/.style={sibling distance=1cm, level distance=1cm},
%     visible on=<3->
%     ]
%   \node[treenode] (tree) {$\otimes$}
%     child { node[treenode] (a) {${\color{blue} a}$} }
%     child { node[treenode] (b){${\color{magenta} \mathsf{inl}}$}
%         child { node[treenode] (c) {${\color{magenta} b}$} }
%       };
%       \node[draw, fit=(tree)(a)(b)(c),
%       label={[name=l] $g(ab)$}, visible on=<3->] {};
%     \node[below=2.5cm of tree] (mt) {};
%     \node[below=2.5cm of a] (mta) {};
%     \node[below=2.5cm of b] (mtb) {};
%     \node[below=-0.2cm of mt, visible on=<4->] (mtset) {};
%       \node[draw, fit=(mt)(mtset)(mta)(mtb),
%       label={[name=l, visible on=<4->] $g(aa)$}, visible on=<4->] {};
% \end{tikzpicture}

%% 1. First example: finite grammars (regex minus Kleene star)
%%    - why ordered linear:
%%      + no exchange
%%      + no contraction
%%      + no weakening
%%      + ambiguity
%% 2. Second example: Kleene star
%% 3. Third example: NFA traces to show the indexed inductive types

@@

>>>>

% -*- fill-column: 80; -*-
\documentclass[review,anonymous,screen,acmsmall,nonacm]{acmart}
\usepackage{mathpartir}
\usepackage{tikz-cd}
\usepackage{lipsum}
\usepackage{enumitem}
\usepackage{wrapfig}
\usepackage{fancyvrb}
\usepackage{xspace}
\usepackage{listings}
\usepackage{hyperref}
\usepackage{cleveref}
\usepackage{stmaryrd}
\usepackage{scalerel}
\usepackage{tikz}
\usetikzlibrary{automata, positioning, arrows, fit}

@@

\newcommand{\bang}{~\textbf{!}~}
\newcommand{\nil}{\mathsf{nil}}
\newcommand{\cons}{\mathsf{cons}}
\newcommand{\data}{\mathsf{data}}
\newcommand{\where}{\mathsf{where}}
\newcommand{\Trace}{\mathsf{Trace}}
\newcommand{\stringquote}[1]{\texttt{\textquotesingle#1\textquotesingle}}

@@

\steven{To save space, each of these example figures can be
  minified/potentially inlined}

To gain intuiton for working in \theoryname, let's walk through some examples of
grammars and parses internal to it. In each of our examples, we will look at
grammars over the three character alphabet $\{ a , b , c\}$.

Each letter of the
alphabet corresponds to a base type in our calculus. For instance, the grammar
for the single character $a$ has a single parse tree for the string \stringquote{a}.
and no parse trees at any other strings.

\paragraph{Finite Grammars}
First consider finite grammars --- those built from base types via dijunctions $\oplus$ and
concatenations $\otimes$. For instance, $(a \oplus b) \otimes c$.
In \cref{fig:fingram} we find a proof tree showing that the input string \stringquote{ac} matches
the grammar $(a \oplus b) \otimes c$.

@@

\paragraph{Regular Expressions}
Next, we demonstrate how to parse a regular expression in \theoryabbv. Similar
to finite grammars, regular
expresssions are built inductively using base types, $\oplus$, and $\otimes$ ;
however, unlike finite grammars, regular expressions additionally have a
restricted form of recursion via Kleene star $(\cdot)^{\ast}$.

% TODO this should be beter formatted but can't use listings with \otimes ^{\ast}
\begin{figure}
\begin{align*}
\data &~A^{*}: L_{i}~\where\\
      & \nil : I \lto A^{*} \\
      & \cons : A \otimes A^{*} \lto A^{*}
\end{align*}
\caption{Kleene Star as an inductive type}
\label{fig:kleenestarinductive}
\end{figure}

@@

However, we can have derivations where the term in context is not simply a
string of base types. For instance, by using the elimination principle for
Kleene star, \cref{fig:kleeneabstractproof} builds a term of type
$A^{*}$ in context comprising abstract linear types. In these contexts, it is
not quite appropriate
to treat linear terms as parse \textit{trees}, because the information in
context is not a string. Instead, it is more
appropriate to think of such terms as parse \textit{transformers}. That is, in
\cref{fig:kleeneabstractproof} the
inhabitant $x : (A \otimes A)^{*}$ in the context is not itself data to be parsed; rather,
$x$ is a parse tree and the term from context $x : (A \otimes A)^{*}$ to the
type $A^{*}$ is a transformation of parse trees of the input grammar into parse
trees of the output grammar.

@@

\paragraph{Non-deterministic Finite Automata}
As a final example, consider the non-deterministic finite
automaton (NFA) $N$ from \cref{fig:exampleNFA}. Such automata play a critical
role, as they serve as operationalized and
executable presentation of a formal grammar. Therefore, it is crucial that we
can represent them internal to \theoryabbv.

Our method of representing these automata internally is to provide an \emph{indexed}
inductive type of traces through the machine. For $N$, this indexed inductive
type is given in \cref{fig:nfainductive}.

The type of traces in $N$ is indexed by two pieces of data: a state in $N$ and a
Boolean. For each state $s : N.\mathsf{state}$ and $acc : \Bool$, we define the
type of traces from state $s$, $\Trace_{N}~s~acc$. The Boolean flag $acc$ marks if the trace
ends in either an accepting or a rejecting state. Let us dissect the structure
of this type in more detail for state 1 of $N$.

$\Trace_{N}~1~acc$ has three constructors. The first constructor
$\mathsf{stop}_{1}$ allows a trace to end at state one, and because state
1 is accepting $acc$ is true when terminating at state 1.

The constructor
$\mathsf{1to2}$ denotes the $b$-labeled transition in $N$ from state 1 to state
2. The character $b$ is consumed when taking this transition and then a trace
beginning at state 2 must be produced, which is denoted by the tensor product of
the grammar $b$ and $Trace_{N}~2~acc$. Finally, the constructor $\mathsf{1to2}$
abstracts over the acceptance Boolean because it is not yet determined.
Because the Boolean is set by the acceptance criteria of the \emph{final} state,
it is provided inductively by the trace out of state 2.

The constructor $\mathsf{1to3}$ is built quite similarly as $\mathsf{1to2}$;
however, $\mathsf{1to3}$ does not involve a tensor product. This is because the
transition from state 1 to state 3 is labeled by the empty string $\epsilon$, so
the transition consumes no input. In this case, when building a trace out of
state 1, it suffices to build a trace out of state 3 \emph{over the same string}.

\begin{figure}
  \begin{tikzpicture}[node distance = 25mm ]
    \node[state, initial, accepting] (1) {$1$};
    \node[state, below left of=1] (2) {$2$};
    \node[state, right of=2] (3) {$3$};

    \path[->] (1) edge[above] node{$b$} (2)
              (1) edge[below, bend right, left=0.3] node{$\epsilon$} (3)
              (2) edge[loop left] node{$a$} (2)
              (2) edge[below] node{$b$} (3)
              (3) edge[above, bend right, right=0.3] node{$a$} (1);
  \end{tikzpicture}
  \caption{NFA $N$}
  \label{fig:exampleNFA}
\end{figure}

\begin{figure}
\begin{align*}
\data &~\Trace_{N}~1~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{1} : I \lto \Trace_{N}~1~\true \\
      & \mathsf{1to2} : \forall (acc : \Bool) \to b \otimes \Trace_{N}~2~acc \lto \Trace_{N}~1~acc \\
      & \mathsf{1to3} : \forall (acc : \Bool) \to \Trace_{N}~3~acc \lto \Trace_{N}~1~acc \\
\\
\data &~\Trace_{N}~2~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{2} : I \lto \Trace_{N}~2~\false \\
      & \mathsf{2to2} : \forall (acc : \Bool) \to a \otimes \Trace_{N}~2~acc \lto \Trace_{N}~2~acc \\
      & \mathsf{2to3} : \forall (acc : \Bool) \to b \otimes \Trace_{N}~3~acc \lto \Trace_{N}~2~acc \\
\\
\data &~\Trace_{N}~3~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{3} : I \lto \Trace_{N}~3~\true \\
      & \mathsf{3to1} : \forall (acc : \Bool) \to a \otimes \Trace_{N}~1~acc \lto \Trace_{N}~3~acc \\
\end{align*}
\caption{$N$ as an indexed inductive type}
\label{fig:nfainductive}
\end{figure}

%% 3. Third example: NFA traces to show the indexed inductive types

@@

\steven{this entire section should become much more compact with the examples
  section from the beginning in mind}
